Genetic Dilution erodes productivity: Exploring farmers' low adoption levels of improved maize in Ethiopia

### 2022-11-04 document/slides/NEUDC 2022/main.tex
Updates from Overleaf

%\usetheme{Warsaw}
\usetheme{Madrid}
\usecolortheme{default}

@@

\author[et al]{Aleksandr Michuda \\ Cristina Chiarella \\Oscar Barriga-Cabanillas \\Juan Correa }

@@

\date[January 2022] % (optional)
{CGIAR-SPIA \\
January, 2022}

@@

%---------------------------------------------------------
\begin{frame}
\frametitle{Research question}

@@

\begin{enumerate}[ {[}1{]} ]
    \item Market integration
    \item Access to supplementary inputs
    \item Agro-ecological considerations
\end{enumerate}

\textbf{We study how heterogeneity masks the true impact of adoption}

\begin{enumerate}[ {[}1{]} ]
    \item Explore heterogeneous comparative advantages through GRC: find puzzling no significant results.
    \item Compare definitions of improved seeds: self-reported, second generation or recycled, purity of genetic material, year of release, types and source.
    \item Low adoption of newer and purest varieties, which do return in higher yields. Poor farmers paying a premium for genetically diluted improved varieties.
\end{enumerate}

@@

%---------------------------------------------------------
\begin{frame}
\frametitle{Data}

\begin{itemize}
    % \item Struggled to provide an adequate and reliable food supply
    \item We focus on maize producing households 
    \item Self-reported use of improved seed varieties
    \item Compare self-reported and cropcut yields
\end{itemize}

@@

\textbf{Understand the returns of adopting improved varieties: }
\begin{itemize}
    \item Some farmers have higher returns to adoption (comparative advantage)
    \item How comparative advantage plays a role on using improved seed varieties.
\end{itemize}

We use a Correlated Random Coefficient Model:
\vspace{-2em}
\begin{center}
$$
    y_{it}= \delta + \Delta h_{it} + \theta_{i} + \phi\theta h_{it} + \tau_{i} + \epsilon_{it}
$$  
\end{center}

\begin{itemize}
    \item $y_{it}$: yields or profits for household i at time t
    \item $h$:  indicator for adoption
    \item $\tau$:  household’s absolute advantage
    \item \textbf{$\theta$: household’s comparative advantage}
    \item \textbf{$\Delta$: The marginal effect of adoption}
    \item $\phi$: Gains from adopting, relative to a household’s comparative advantage.
\end{itemize}

@@

Advantages:
\begin{enumerate}
    \item Identification does not rely on a IV 
    % (linear projection of observed adoption history)
    \item Disentangles absolute advantage from comparative advantage in adoption. ($\theta_i = \theta_A - \theta_N$)
\end{enumerate}

\input{results/tables/traj}

% Extending the model:

% \begin{itemize}
%     \item Implementing a nonparametric panel identification in a novel GRC (Group Random Coefficients)
%     \item Modeling and incorporating time-varying characteristics 
%     \item Complement ESS data with precipitation (CHIRPS) and temperature (CPC) data
% \end{itemize}  
\end{frame}

%---------------------------------------------------------
%---------------------------------------------------------
\begin{frame}
\frametitle{Empirical Strategy II: Returns to Different Types of Hybrid}
% \frametitle{Group Random Coefficients Results}
\textbf{Why farmers do \textit{(not)} adopt?}

\begin{enumerate}
    \item There are negative returns from adoption
    \item Comparative advantage is close to 0 unless you always adopt (perhaps driven by 
\end{enumerate}

\begin{figure}
    \centering
    \label{fig:traj_sum}
    \includegraphics[scale=.4]{results/figures/traj_sum.png}
\end{figure}

@@

Why do we see farmers adopting despite the seemingly small negative returns?

\begin{itemize}
    \item We use the DNA-fingerprinting to shed light on this puzzle
    \item DNA allows us to assess what varieties (and purity) farmers are using 
    \item Unfortunately, only available in the last round
\end{itemize}

We implement a switching regression approach

@@

\input{results/tables/switch23-sr}

@@

%---------------------------------------------------------
\begin{frame}
\frametitle{Discussion}
Negative and small returns to adoption from self-reported improved varieties.
\begin{itemize}
    \item But positive returns to more genetically pure, DTMZ, CG- sourced and newer varieties.
    \item All DNA fingerprints are 70\% pure or higher. Control group has high share of false positives: no proper control.

\end{itemize}
Negative GRC results mask a large heterogeneity in returns to adoption. 
\begin{itemize}
    \item Low purchase of newly bred hybrids, diluted DNA, low inputs use.

\end{itemize}

Effects of high-yielding varieties diminishes over time, but traditional varieties are well adapted to local agro-ecological conditions.

\begin{itemize}
    \item There may be a threshold in time above which it may be more profitable to use a traditional variety than an improved one that has been around for too many generations.

\end{itemize}

@@

%---------------------------------------------------------
\begin{frame}{}
    Thank you!
\end{frame}

\begin{frame}{Raw Trajectories}
    \begin{figure}
        \centering
        \includegraphics[scale=.5]{results/figures/theta.png}
    \end{figure}
\end{frame}

>>>>

%\usetheme{Warsaw}
\usetheme{Madrid}
\usecolortheme{default}
\usepackage{natbib}
\bibliographystyle{unsrtnat}

@@

\author[et al]{Aleksandr Michuda\inst{1} \and Cristina Chiarella\inst{2} \and Oscar Barriga-Cabanillas\inst{3} \and Juan Correa\inst{4}}

\institute{\inst{1} Cornell University \and \inst{2} UCLouvain \and \inst{3} World Bank Corp. \and \inst{4} FAO}

@@

\date[November 2022] % (optional)
{NEUDC   \\
November 2022}

@@

%---------------------------------------------------------
\begin{frame}
\frametitle{Introduction}

@@

\begin{enumerate}[ {[}1{]} ]
    \item Market integration
    \item Access to supplementary inputs
    \item Agro-ecological considerations 
    \item Lack of learning \citep{Conley2010-ue}
\end{enumerate}

@@

\begin{frame}{Research Question}

\textbf{How does misclassification in adoption status mask heterogeneous impacts to adoption?}
\end{frame}

\begin{frame}{Preview of Results}

\begin{enumerate}[ {[}1{]} ]
    \item Explore heterogeneous comparative advantages through GRC: find puzzling results.
    \item Endogenous switching regression with different definitions of improved seeds: self-reported, second generation or recycled, purity of genetic material, year of release, types and source.
    \item Low adoption of newest and purest varieties, which do return in higher yields. 
    \item When we account for misclassification in GRC, the puzzle is resolved
\end{enumerate}
\end{frame}

%---------------------------------------------------------
\begin{frame}
\frametitle{Ethiopian Maize}

\begin{itemize}
    % \item Struggled to provide an adequate and reliable food supply
    \item Use ESS waves 1-4, agricultural questionnaires
    \item Waves 1-3 were a panel and included information on yields, inputs and hybrid maize adoption
    \item Wave 4 is a break from Waves 1-3, but contains similar information $+$ DNA fingerprinting
    \item We focus on maize producing households 
    \item Self-reported use of improved seed varieties
    \item Self-reported yields
\end{itemize}

@@

\begin{frame}{}

\begin{figure}
    \centering
    \includegraphics[width=.7\textwidth]{results/figures/adoption_r4.png}
    \caption{Caption}
    \label{fig:my_label}
\end{figure}

@@

\textbf{Understand the returns of adopting improved varieties: }
\begin{itemize}
    \item Some farmers are ``more suited'' for adoption (comparative advantage) \citep{Suri2011-oi}
    \item How comparative advantage plays a role on using improved seed varieties and its returns.
    \item Use Group Random Coefficient (GrRC) method \citep{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}
    \item Can separate returns from those who ``should'' be getting highest returns.
\end{itemize}

We estimate a Correlated Random Coefficient Model by way of GrRC estimation:
\vspace{-2em}
\begin{center}
$$
    y_{it}= \delta + \Delta h_{it} + \phi\boldsymbol{\theta_i} h_{it} + \tau_{i} + \boldsymbol{\theta_i} + \epsilon_{it}
$$  
\end{center}

%---------------------------------------------------------
\begin{frame}
\frametitle{Empirical Strategy}
\begin{enumerate}
    \item Identification does not rely on an IV or propensity to adopt.
    \item Identification similar to \cite{Chamberlain1984-uk}: strict exogeneity
    \item strong assumption, but gives us ability to estimate returns to adoption, as well as \textit{advantage in adopting}
    \item Requires enough variation in ``switcher trajectories'' 
\end{enumerate}

\input{results/tables/traj}
\end{frame}

%---------------------------------------------------------
%---------------------------------------------------------
\begin{frame}
\frametitle{The Puzzle}
% \frametitle{Group Random Coefficients Results}

\begin{figure}
    \centering
    \includegraphics[scale=0.5]{results/figures/theta.png}\label{fig:theta_delta_raw}
\end{figure}

@@

%---------------------------------------------------------
\begin{frame}
\frametitle{What's going on?}

\begin{itemize}
    \item Negative advantage to adopting
    \item Adopting more associated with less advantage
    \item Negative returns (but that's part of the standard story about access)
\end{itemize}

\begin{frame}{How do we resolve this puzzle?}
\begin{itemize}
    \item Explore 4th wave
    \item We use the DNA-fingerprinting to shed light on this puzzle
    \item DNA allows us to assess what varieties (and purity) farmers are using, 
    \item We implement an endogenous switching regression approach
\end{itemize}

@@

\begin{table}[H]
\resizebox{.75\textwidth}{!}{
\centering
\hspace*{-1.2cm}
\begin{threeparttable}
\caption{Effects on yields (log) from adopting improved maize varieties (DNA fingerprinting)}
\label{tab:switch2}
\begin{tabular}{l cccccc}
\hline
\hline
            &Adopters yields&Non-adopters yields&         ATE&          SE&     p-value\\
\hline
\textit{DNA 70, SR yields}&            &            &            &            &            \\
ATT         &        7.17&        7.31&      \fcolorbox{red}{lightgray}{-0.14}&       0.026&      0.0000\\
%
%
%
ATU         &        8.76&        6.76&        2.00&       0.036&      0.0000\\
%
%
%
\textit{DNA 90, SR yields}&            &            &            &            &            \\
ATT         &        7.18&        9.07&       \fcolorbox{red}{lightgray}{-1.90}&       0.023&      0.0000\\
%
%
%
ATU         &        7.01&        6.80&        0.21&       0.022&      0.0000\\
%
%
%
\textit{DNA 95, SR yields}&            &            &            &            &            \\
ATT         &        7.30&        7.08&        \fcolorbox{red}{lightgray}{0.22}&       0.036&      0.0000\\
%
%
%
ATU         &        6.87&        6.86&       0.018&       0.029&      0.5496\\
%
%
%
\textit{HYB, SR yields}&            &            &            &            &            \\
ATT         &        7.28&        7.25&       \fcolorbox{red}{lightgray}{0.030}&       0.028&      0.2892\\
%
%
%
ATU         &        7.06&        6.85&        0.21&       0.028&      0.0000\\
%
%
%
\textit{DTMZ, SR yields}&            &            &            &            &            \\
ATT         &        7.42&        6.53&        \fcolorbox{red}{lightgray}{0.89}&       0.074&      0.0000\\
%
%
%
ATU         &        7.11&        7.02&       0.089&       0.022&      0.0001\\
%
%
%
\textit{CGIAR source, SR yields}&            &            &            &            &            \\
ATT         &        7.13&        6.71&        \fcolorbox{red}{lightgray}{0.43}&       0.033&      0.0000\\
%
%
%
ATU         &        9.19&        6.89&        2.31&       0.018&      0.0000\\
%
%
%
\textit{Year 2010+, SR yields}&            &            &            &            &            \\
ATT         &        7.44&        6.92&        \fcolorbox{red}{lightgray}{0.53}&       0.039&      0.0000\\
%
%
%
ATU         &        8.98&        6.91&        2.07&       0.056&      0.0000\\
%
%
%
\hline
\hline
\end{tabular}
\end{threeparttable}
}
\end{table}

@@

%---------------------------------------------------------
\begin{frame}
\frametitle{Discussion}
Negative and small returns to adoption from self-reported improved varieties.
\begin{itemize}
    \item But positive returns to more genetically pure, DTMZ, CG- sourced and newer varieties.
    \item All DNA fingerprints are 70\% pure or higher. Control group has high share of false positives: no proper  control.  
\end{itemize}

Could misclassification be masking returns heterogeneity?
\begin{itemize}
    \item If farmers using some mix of traditional hybrid maize or using incorrect practices, this would attenuate effects
    \item Trajectories would be misclassified.
\end{itemize}

@@

\begin{frame}{Misclassification Driving Returns?}
\begin{itemize}
    \item Use a misclassification robust variant of GRC
    \item Maximum likelihood estimation that uses misclassification rates in estimation, inspired by \cite{michuda2021three}
    \item Use Wave 4 misclassification rates for Waves 1-3
\end{itemize}

\begin{equation*}
\label{eq:mr-grc}
l(\mu, \kappa, \Delta, \beta, \sigma|y_{t}, \hat{h}) = \sum_{\underbar{$h$}\in\mathcal{H}} \left[ f_{Y_{t}| \mathcal{H}}(y|h=\underbar{$h$})\cdot  \sum_{\bar{h} \in \mathcal{H}} \boldsymbol{1\{\hat{h}=\bar{h}\} p_{\underbar{$h$}, \bar{h}}} \right]
\end{equation*}

\begin{frame}{Misclassification Rates from Wave 4}

\begin{itemize}
    \item Use rate of reported hybrid of at least 95\% purity, against genetic fingerprinting to calculate misclassification
    \item Farmer who reports using traditional seed actually using hybrid 14\% of the time
    \item Farmer who reports using hybrid seed is actually using traditional seed almost \textit{30\%} of the time.
\end{itemize}
    \begin{table}[]
        \centering
        \begin{tabular}{lll}
        \toprule
         & 0 & 1 \\
        \midrule
        0 & 0.863 & 0.137 \\
        1 & 0.296 & 0.704 \\
        \bottomrule
        \end{tabular}
    \end{table}

\begin{equation}
    \label{eq:misclass}
    P(h = (i,j,k) , \hat{h} = (\hat{i},\hat{j}, \hat{k})) = p_{i, \hat{i}}\cdot p_{j, \hat{j}}\cdot p_{k, \hat{k}}
\end{equation}
\end{frame}

\begin{frame}{Misclassified GRC Results}

\begin{figure}
    \centering
    \includegraphics[scale=0.4]{results/figures/mr-theta.png}
\end{figure}
\end{frame}

\begin{frame}{Conclusion}

Accounting for misclassification resolves the issue and makes the adoption problem the classic story about access and a lack of profitability.

\begin{itemize}
    \item Misclassification may work through a measurement problem, but also through changes in practices
    \item Mismatch between cultivation practices and seed used
    \item Effects of high-yielding varieties diminish over time, but traditional varieties are well adapted to local agro-ecological conditions.
    \begin{itemize}
    \item There may be a threshold in time above which it may be more profitable to use a traditional variety than an improved one that has been around for too many generations.
    \item Example: 010 trajectory
\end{itemize}
\end{itemize}

\begin{frame}
Thank you!

\begin{itemize}
    \item I would be happy to discuss any questions and please feel free to email me!
    \item am2497@cornell.edu
\end{itemize}
